---
title: "run_analysis"
output: html_document
---
This markdown file explains a code that distills a "tidy_data_set.txt" out of a zip file with several datasets from Human Activity Recognition using a Smartphones Dataset. 

First create a variable with the URL as a link to the data (URL) in your working environement.
```{r}
URL <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
```
Set up the sample data file, the working directory, load the libraries and clean the global environement.
```{r}
run_analysis <- function(URL)
         if (!getwd() == "./sample_data") {     # Look for working directory and check if file "sample data" exists otherwise create it
                dir.create("./sample_data")}
setwd("./sample_data") # Set working directory
rm(list = ls(all = TRUE)) # Clear the global environement within working directory
library(plyr) # Load plyr package
library(data.table) # Load data.table package (fast data handling)
library(dplyr) # Load dplyr package (newer brother of plyr)
```
Set up a temporary file, read out the content of the .zip file and read out manually selected files
```{r}
tmp <- tempfile() # Create temporay file for the download file
download.file("https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip",tmp,method = "curl") # Download file into tmp
unzip(tmp, list = TRUE) # Unzip temp to extract the list of content (to decide which data is applicable) but not the data yet
YTest <- read.table(unzip(tmp, "UCI HAR Dataset/test/y_test.txt")) # Test set
XTest <- read.table(unzip(tmp, "UCI HAR Dataset/test/X_test.txt")) # Test labels
SubjectTest <- read.table(unzip(tmp, "UCI HAR Dataset/test/subject_test.txt")) # Identifier for the participants performing the test (-activity)
YTrain <- read.table(unzip(tmp, "UCI HAR Dataset/train/y_train.txt")) # Training set
XTrain <- read.table(unzip(tmp, "UCI HAR Dataset/train/X_train.txt")) # Training labels
SubjectTrain <- read.table(unzip(tmp, "UCI HAR Dataset/train/subject_train.txt")) # Identifier for the subjects performing the training (-activity)
Features <- read.table(unzip(tmp, "UCI HAR Dataset/features.txt")) # Selected and filtered features (e.g. tBodyAcc-XYZ...) from a 3D raw data signal database
unlink(tmp) # Remove link to the temporary file (tmp) and free memory
```       
Insert column names (variables) into the data sets (XTrain, XTest)
```{r}
colnames(XTrain) <- t(Features[2])# Use t() to transpose all features from column to row and add them into the data sets
colnames(XTest) <- t(Features[2])
```        
Add to the XTest and XTrain data set activity and participants variables
```{r}
XTrain$activities <- YTrain[, 1] # Add column with activity labels to XTrain, now 562 variables
XTrain$participants <- SubjectTrain[, 1] # Add participants variable
XTest$activities <- YTest[, 1] # Add activities variable, now 562 variables
XTest$participants <- SubjectTest[, 1] # Add participants variable
```
Merges X-Training and X-Test set and remove duplicates
```{r}
Master <- rbind(XTrain, XTest) 
duplicated(colnames(Master)) # Find duplicates in the variables (package:data.table / overall finding = 84 duplicates)
Master <- Master[,!duplicated(colnames(Master))] # Generate master without duplicates
```
Extracts only the measurements on the mean and standard deviation (std) for each measurement and create a data.frame for the mean and std measuerments
```{r}
Mean <- grep("mean()", names(Master), value = FALSE, fixed = TRUE) # Extract all features with pattern "mean()" from Master
# In addition, we need to include 555:559 as they have means and are associated with the gravity terms
Mean <- append(Mean, 471:477) # Add further featuers with mean measurements (new rows in Mean 471:477)
FeatureMeanMatrix <- Master[Mean] # Data.frame with features only including mean measurements
# For Standard Deviation (STD) the same procedure: Set up a data.frame with std measurements
STD <- grep("std()", names(Master), value = FALSE) # Extract all features with pattern "std()"
 # Create data.frame with freatuere only including the std measurements
FeatureSTDMatrix <- Master[STD]
```
Use descriptive activity names (as described in activitylabels.txt) to rename the activities #in the data.frame "Master"
```{r}
# Change data typ to "character"
Master$activities[Master$activities == 1] <- "Walking"
Master$activities[Master$activities == 2] <- "Walking Upstairs"
Master$activities[Master$activities == 3] <- "Walking Downstairs"
Master$activities[Master$activities == 4] <- "Sitting"
Master$activities[Master$activities == 5] <- "Standing"
Master$activities[Master$activities == 6] <- "Laying"
Master$activities <- as.factor(Master$activities) # Change data typ to "factor"
```
Appropriately label the data set with descriptive variable names (e.g. "Acc"="Accelerator")
```{r}
names(Master)  # Survey all names in "Master"
names(Master) <- gsub("Acc", "Accelerator", names(Master))
names(Master) <- gsub("Mag", "Magnitude", names(Master))
names(Master) <- gsub("Gyro", "Gyroscope", names(Master))
names(Master) <- gsub("^t", "time", names(Master))
names(Master) <- gsub("^f", "frequency", names(Master))
```
Label the participants numbers appropriately (e.g. "1" = "Participant 1")
```{r}
Master$participants <- as.character(Master$participants) # Change data typ to "character" 
Master$participants[Master$participants == 1] <- "Participant 1"
Master$participants[Master$participants == 2] <- "Participant 2"
Master$participants[Master$participants == 3] <- "Participant 3"
Master$participants[Master$participants == 4] <- "Participant 4"
Master$participants[Master$participants == 5] <- "Participant 5"
Master$participants[Master$participants == 6] <- "Participant 6"
Master$participants[Master$participants == 7] <- "Participant 7"
Master$participants[Master$participants == 8] <- "Participant 8"
Master$participants[Master$participants == 9] <- "Participant 9"
Master$participants[Master$participants == 10] <- "Participant 10"
Master$participants[Master$participants == 11] <- "Participant 11"
Master$participants[Master$participants == 12] <- "Participant 12"
Master$participants[Master$participants == 13] <- "Participant 13"
Master$participants[Master$participants == 14] <- "Participant 14"
Master$participants[Master$participants == 15] <- "Participant 15"
Master$participants[Master$participants == 16] <- "Participant 16"
Master$participants[Master$participants == 17] <- "Participant 17"
Master$participants[Master$participants == 18] <- "Participant 18"
Master$participants[Master$participants == 19] <- "Participant 19"
Master$participants[Master$participants == 20] <- "Participant 20"
Master$participants[Master$participants == 21] <- "Participant 21"
Master$participants[Master$participants == 22] <- "Participant 22"
Master$participants[Master$participants == 23] <- "Participant 23"
Master$participants[Master$participants == 24] <- "Participant 24"
Master$participants[Master$participants == 25] <- "Participant 25"
Master$participants[Master$participants == 26] <- "Participant 26"
Master$participants[Master$participants == 27] <- "Participant 27"
Master$participants[Master$participants == 28] <- "Participant 28"
Master$participants[Master$participants == 29] <- "Participant 29"
Master$participants[Master$participants == 30] <- "Participant 30"
Master$participants <- as.factor(Master$participants) # Change data typ to "factor"
```
Create the final tidy data set
```{r}
Master.dt <- data.table(Master) # Change data.frame to data.table (data.table package)
TidyData <- Master.dt[, lapply(.SD, mean), by = 'participants,activities']
#This takes the mean of every column broken down by participants and activities
write.table(TidyData, file = "tidy_data_set.txt", row.names = FALSE) # Write data.table into .txt in file "sample_data"
```


